---
title: "Session 2: PCOS and Inflammatory Markers in Menstrual Cycle"
author: "Data Science for Biology Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
  pdf:
    toc: true
    toc-depth: 3
  docx:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
editor: visual
---

```{r setup, include=FALSE}
# ============================================================================
# Setup: Load packages, set options
# ============================================================================

library(tidyverse)
library(gtsummary)
library(ggpubr)
library(lmerTest)
library(flextable)
library(officer)
library(knitr)

# Set seed for reproducibility
set.seed(42)

# Output directory
output_dir <- "."
```

## Research Question

> **Do inflammatory marker levels differ between participants with PCOS and participants without PCOS after accounting for menstrual-cycle stage?**

In this analysis we compare inflammatory cytokine concentrations between participants diagnosed with PCOS and those with no disease, while adjusting for the time point (menstrual-cycle stage) as a covariate.

------------------------------------------------------------------------

## Data Import

```{r data-import}
# ============================================================================
# Read all CSV files from the dataset directory
# ============================================================================

# Participant metadata: age, PCOS status, arm, period product, sex
participant_meta <- read_csv("data/group_b_menstruation/01_participant_metadata_period.csv",
                             show_col_types = FALSE)

# Sample IDs mapping: links participants to their samples across time points
sample_ids <- read_csv("data/group_b_menstruation/00_sample_ids_period.csv",
                       show_col_types = FALSE)

# Luminex cytokine concentrations
luminex <- read_csv("data/group_b_menstruation/02_luminex_period.csv",
                    show_col_types = FALSE)

# Preview each dataset
glimpse(participant_meta)
glimpse(sample_ids)
glimpse(luminex)
```

## Data Validation

```{r data-validation}
# ============================================================================
# Validate key identifiers and merge datasets
# ============================================================================

# --- Participant metadata checks ---
cat("Number of participants:", n_distinct(participant_meta$pid), "\n")
cat("PCOS status levels:", unique(participant_meta$pcos_status), "\n")
cat("Sex levels:", unique(participant_meta$sex), "\n")
cat("Arms:", unique(participant_meta$arm), "\n")
cat("Period products:", unique(participant_meta$period_product), "\n")

# --- Sample ID checks ---
cat("\nNumber of samples:", n_distinct(sample_ids$sample_id), "\n")
cat("Time points:", unique(sample_ids$time_point), "\n")
cat("Unique PIDs in sample_ids:", n_distinct(sample_ids$pid), "\n")

# Check that all PIDs in sample_ids exist in participant_meta
pid_in_both <- all(sample_ids$pid %in% participant_meta$pid)
cat("\nAll sample PIDs found in participant metadata:", pid_in_both, "\n")

# --- Luminex checks ---
cat("\nUnique cytokines:", unique(luminex$cytokine), "\n")
cat("Number of cytokines:", n_distinct(luminex$cytokine), "\n")
cat("Unique sample IDs in luminex:", n_distinct(luminex$sample_id), "\n")

# Check that all luminex sample IDs exist in sample_ids
sample_in_both <- all(luminex$sample_id %in% sample_ids$sample_id)
cat("All luminex sample IDs found in sample_ids:", sample_in_both, "\n")

# --- Merge into analysis dataset ---
# Step 1: Join sample_ids with participant_meta
# Both datasets have 'arm', so we keep it from participant_meta
analysis_data <- sample_ids %>%
  left_join(participant_meta, by = c("pid", "arm"))

# Step 2: Join with luminex
analysis_data <- analysis_data %>%
  left_join(luminex, by = "sample_id")

# Step 3: Create log-transformed concentration
# log1p = log(1 + x) handles zeros gracefully
analysis_data <- analysis_data %>%
  mutate(log_conc = log1p(conc))

# Final dataset: select the columns we need
analysis_data <- analysis_data %>%
  select(pid, pcos_status, age, arm, period_product, time_point,
         cytokine, conc, log_conc)

# Remove any rows with missing cytokines (no luminex measurement)
analysis_data <- analysis_data %>%
  filter(!is.na(cytokine))

cat("\nFinal analysis dataset dimensions:", dim(analysis_data), "\n")
cat("Number of participants:", n_distinct(analysis_data$pid), "\n")
cat("Number of observations:", nrow(analysis_data), "\n")

glimpse(analysis_data)
```

------------------------------------------------------------------------

## Table 1: Participant Characteristics by PCOS Status

```{r table1}
# ============================================================================
# Create Table 1 using gtsummary, grouped by pcos_status
# ============================================================================

# Use distinct participant-level data (one row per participant)
participant_tbl <- participant_meta %>%
  distinct(pid, .keep_all = TRUE)

# Remove 'sex' if it has only one level — it does (all "female")
sex_levels <- unique(participant_tbl$sex)
cat("Sex levels:", sex_levels, "\n")

# Build Table 1
tbl1 <- participant_tbl %>%
  select(pcos_status, age, arm, period_product) %>%
  tbl_summary(
    by = pcos_status,
    label = list(
      age ~ "Age (years)",
      arm ~ "Birth control use",
      period_product ~ "Period product"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(age ~ c(1, 1))
  ) %>%
  add_p() %>%
  add_overall()

# Display in document
tbl1

# Export to HTML
tbl1 %>%
  as_flex_table() %>%
  flextable::save_as_html(path = "Table_1_PCOS_vs_NoDisease.html")

# Export to Word
tbl1 %>%
  as_flex_table() %>%
  flextable::save_as_docx(path = "Table_1_PCOS_vs_NoDisease.docx")

cat("Table 1 saved to Table_1_PCOS_vs_NoDisease.html and .docx\n")
```

------------------------------------------------------------------------

## Cytokine Comparison Figure

```{r cytokine-figure, fig.width=12, fig.height=10}
# ============================================================================
# Boxplots of log-transformed cytokine concentrations by PCOS status
# ============================================================================

# Detect cytokines dynamically from the data
cytokines <- unique(analysis_data$cytokine)
cat("Cytokines detected:", paste(cytokines, collapse = ", "), "\n")

# Create the boxplot
p <- analysis_data %>%
  ggboxplot(x = "pcos_status", y = "log_conc",
            color = "pcos_status",
            palette = c("#E69F00", "#56B4E9"),
            add = "jitter",
            add.params = list(alpha = 0.4, size = 1.2)) +
  stat_compare_means(method = "wilcox.test",
                     label = "p.signif",
                     hide.ns = FALSE,
                     size = 3.5) +
  facet_wrap(~ cytokine, scales = "free_y", ncol = 4) +
  labs(title = "Inflammatory marker levels in PCOS and non-PCOS participants",
       x = "PCOS Status",
       y = "log(1 + concentration)") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "bottom"
  )

print(p)

# Save as PNG
ggsave("Figure_1_Cytokine_Comparison.png", plot = p,
       width = 12, height = 10, dpi = 300)

# Save as PDF
ggsave("Figure_1_Cytokine_Comparison.pdf", plot = p,
       width = 12, height = 10)

cat("Figure saved to Figure_1_Cytokine_Comparison.png and .pdf\n")
```

------------------------------------------------------------------------

## Mixed-Effects Model

```{r mixed-model}
# ============================================================================
# Fit a linear mixed-effects model for each cytokine
# Formula: log_conc ~ pcos_status + time_point + (1 | pid)
#   - pcos_status: fixed effect of interest (PCOS vs no disease)
#   - time_point: adjustment for menstrual-cycle stage
#   - (1 | pid): random intercept per participant (accounts for repeated measures)
# ============================================================================

# Get unique cytokines
cytokines <- unique(analysis_data$cytokine)

# Create a results dataframe
model_results <- data.frame()

for (cyt in cytokines) {
  # Subset data for this cytokine
  sub_data <- analysis_data %>%
    filter(cytokine == cyt) %>%
    mutate(
      # Ensure pcos_status is a factor with "no disease" as reference
      pcos_status = factor(pcos_status, levels = c("no disease", "pcos")),
      # Ensure time_point is a factor for the model
      time_point = factor(time_point, levels = c("week_prior", "onset",
                                                  "end_bleeding", "week_post"))
    )

  # Fit the mixed-effects model
  # Using lmerTest which provides p-values via Satterthwaite approximation
  fit <- lmer(log_conc ~ pcos_status + time_point + (1 | pid),
              data = sub_data)

  # Extract the coefficient for pcos_status (PCOS vs no disease)
  coef_summary <- summary(fit)$coefficients

  # Find the row for pcos_status (it will be "pcos_statuspcos")
  pcos_row <- grep("pcos_status", rownames(coef_summary))

  # Get confidence interval
  ci <- confint(fit, parm = "pcos_statuspcos", method = "Wald")

  # Store results
  row_result <- data.frame(
    cytokine = cyt,
    estimate = coef_summary[pcos_row, "Estimate"],
    std_error = coef_summary[pcos_row, "Std. Error"],
    ci_lower = ci[1],
    ci_upper = ci[2],
    p_value = coef_summary[pcos_row, "Pr(>|t|)"],
    n_obs = nrow(sub_data),
    n_participants = n_distinct(sub_data$pid),
    stringsAsFactors = FALSE
  )

  model_results <- rbind(model_results, row_result)
}

# Apply Benjamini-Hochberg correction across all cytokines
model_results$p_adjusted <- p.adjust(model_results$p_value, method = "BH")

# Sort by adjusted p-value
model_results <- model_results %>%
  arrange(p_adjusted)

# Display results
kable(model_results, digits = 4,
      caption = "Mixed-effects model results: PCOS effect on log cytokine concentration")

# Write to CSV
write_csv(model_results, "Model_Results_PCOS_Cytokines.csv")
cat("Model results saved to Model_Results_PCOS_Cytokines.csv\n")
```

## Model Results Summary

```{r model-summary}
# ============================================================================
# Highlight which cytokines are significant after BH correction
# ============================================================================

cat("=== Cytokines with significant PCOS effect (BH-adjusted p < 0.05) ===\n")
significant <- model_results %>% filter(p_adjusted < 0.05)
if (nrow(significant) > 0) {
  print(significant %>% select(cytokine, estimate, p_value, p_adjusted))
} else {
  cat("No cytokines reached statistical significance after BH correction.\n")
}

cat("\n=== All cytokines ranked by adjusted p-value ===\n")
print(model_results %>% select(cytokine, estimate, ci_lower, ci_upper, p_value, p_adjusted))
```

------------------------------------------------------------------------

## Regression Table

```{r regression-table}
# ============================================================================
# Create a regression table using tbl_regression() for the first cytokine
# as a representative model, then build a combined table showing PCOS
# coefficients across all cytokines
# ============================================================================

# Fit a model on the complete dataset (all cytokines) to show as a single
# regression table using the first cytokine as a detailed example
first_cyt <- cytokines[1]
sub_data_first <- analysis_data %>%
  filter(cytokine == first_cyt) %>%
  mutate(
    pcos_status = factor(pcos_status, levels = c("no disease", "pcos")),
    time_point = factor(time_point, levels = c("week_prior", "onset",
                                                "end_bleeding", "week_post"))
  )

fit_first <- lmer(log_conc ~ pcos_status + time_point + (1 | pid),
                  data = sub_data_first)

# Create regression table with tbl_regression
reg_table <- fit_first %>%
  tbl_regression(
    label = list(
      pcos_status ~ "PCOS status",
      time_point ~ "Time point"
    ),
    estimate_fun = ~ style_number(.x, digits = 3),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  add_global_p() %>%
  bold_p()

# Display
reg_table

# Export as Word document
reg_table %>%
  as_flex_table() %>%
  save_as_docx(path = "Model_Table_PCOS.docx")

cat("Regression table saved to Model_Table_PCOS.docx\n")
```

### Cross-Cytokine PCOS Coefficient Summary

```{r cross-cytokine-table}
# ============================================================================
# Build a combined table showing the PCOS coefficient for each cytokine
# ============================================================================

# Create a flextable from model results for reporting
cross_table <- model_results %>%
  mutate(
    across(c(estimate, std_error, ci_lower, ci_upper), ~ round(.x, 3)),
    across(c(p_value, p_adjusted), ~ round(.x, 4)),
    significant = ifelse(p_adjusted < 0.05, "Yes", "No"),
    ci = paste0("(", ci_lower, ", ", ci_upper, ")")
  ) %>%
  select(cytokine, estimate, std_error, ci, p_value, p_adjusted, significant) %>%
  flextable() %>%
  set_header_labels(
    cytokine = "Cytokine",
    estimate = "Estimate",
    std_error = "SE",
    ci = "95% CI",
    p_value = "p-value",
    p_adjusted = "BH-adjusted p",
    significant = "Significant"
  ) %>%
  theme_booktabs() %>%
  autofit()

cross_table
```

------------------------------------------------------------------------

## Results Interpretation

```{r interpretation}
# ============================================================================
# Automatically generate interpretation based on model results
# ============================================================================

n_participants <- n_distinct(analysis_data$pid)
n_pcos <- n_distinct(analysis_data$pid[analysis_data$pcos_status == "pcos"])
n_nodisease <- n_distinct(analysis_data$pid[analysis_data$pcos_status == "no disease"])
n_samples_total <- nrow(analysis_data) / length(cytokines)  # per cytokine
n_cytokines <- length(cytokines)

cat("============================================================\n")
cat("             SESSION 2 INTERPRETATION\n")
cat("============================================================\n\n")

cat("Research Question:\n")
cat("Do inflammatory marker levels differ between participants with PCOS\n")
cat("and participants without PCOS after accounting for menstrual-cycle stage?\n\n")

cat("Sample Sizes:\n")
cat("  Total participants:", n_participants, "\n")
cat("    PCOS:", n_pcos, "\n")
cat("    No disease:", n_nodisease, "\n")
cat("  Total samples per cytokine:", n_samples_total, "\n")
cat("  Cytokines analyzed:", n_cytokines, "\n\n")

cat("Key Findings:\n\n")

sig_cytokines <- model_results %>%
  filter(p_adjusted < 0.05)

if (nrow(sig_cytokines) == 0) {
  cat("After fitting a linear mixed-effects model (log_conc ~ pcos_status + time_point + (1|pid))\n")
  cat("with Benjamini-Hochberg correction for multiple testing across", n_cytokines, "cytokines,\n")
  cat("none of the inflammatory markers showed a statistically significant difference\n")
  cat("between PCOS and non-PCOS participants at the adjusted significance threshold (alpha = 0.05).\n\n")

  # Show the closest to significance
  top_cyt <- model_results %>% slice_min(p_adjusted, n = 1)
  cat("The cytokine closest to significance was", top_cyt$cytokine[1],
      "(BH-adjusted p =", round(top_cyt$p_adjusted[1], 4), ").\n\n")
} else {
  cat("After fitting linear mixed-effects models (log_conc ~ pcos_status + time_point + (1|pid))\n")
  cat("with Benjamini-Hochberg correction for multiple testing, the following cytokine(s)\n")
  cat("showed statistically significant differences between PCOS and non-PCOS participants:\n\n")

  for (i in seq_len(nrow(sig_cytokines))) {
    cyt <- sig_cytokines$cytokine[i]
    est <- round(sig_cytokines$estimate[i], 3)
    ci_l <- round(sig_cytokines$ci_lower[i], 3)
    ci_u <- round(sig_cytokines$ci_upper[i], 3)
    p_adj <- round(sig_cytokines$p_adjusted[i], 4)
    direction <- ifelse(est > 0, "higher", "lower")

    cat("  -", cyt, ": estimate =", est, "(", ci_l, ",", ci_u, "),\n")
    cat("    BH-adjusted p =", p_adj, "\n")
    cat("    Interpretation:", cyt, "levels were", direction, "in PCOS participants\n")
    cat("    compared to non-PCOS participants after adjusting for menstrual-cycle stage.\n\n")
  }
}

cat("Statistical Methods:\n")
cat("  - Linear mixed-effects models with random intercepts per participant\n")
cat("  - Fixed effects: PCOS status (primary predictor), time point (adjustment)\n")
cat("  - Multiple testing correction: Benjamini-Hochberg (BH) procedure\n")
cat("  - Log transformation: log1p (log(1 + concentration))\n\n")

cat("Important Caveat:\n")
cat("  This is an exploratory analysis. The relatively small sample size limits\n")
cat("  statistical power. Results should be interpreted cautiously and validated\n")
cat("  in larger, independent cohorts.\n")
cat("============================================================\n")
```

------------------------------------------------------------------------

## Session Information

```{r session-info}
sessionInfo()
```