---
title: "Data Dictionary — Menstrual Cycle Dataset (Group B)"
subtitle: "What each table means, row by row and column by column"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)

data_dir <- here::here("data", "group_b_menstruation")
```

## Overview

This dataset comes from a study on the **menstrual cycle** (Group B). It contains
four related CSV files that describe samples, participant metadata, cytokine
concentrations measured by Luminex, and immune cell counts measured by flow
cytometry. The tables are linked by the `sample_id` column, and participants are
identified by `pid`.

## 1. `00_sample_ids_period.csv`

```{r}
#| label: load-sample-ids
sample_ids <- read_csv(file.path(data_dir, "00_sample_ids_period.csv"))
glimpse(sample_ids)
```

**What a single row represents:** one biological sample collected from a
participant at a specific time point relative to their menstrual period, giving
it a unique `sample_id`.

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | Participant ID (e.g. `pid_01`, `pid_12`). Links to `01_participant_metadata_period.csv`. |
| `time_point` | character | When the sample was taken relative to the period: `onset` (start of bleeding), `end_bleeding`, `week_prior` (one week before), or `week_post` (one week after). |
| `arm` | character | Study arm: `birth_control` or `no_birth_control`. |
| `sample_id` | character | **Unique** sample identifier (e.g. `SAMP001`). Used to link to the Luminex and flow cytometry tables. |

This is the **linking table**: it connects sample IDs to participants and time
points. Each participant may appear multiple times (once for each time point at
which they provided a sample).

---

## 2. `01_participant_metadata_period.csv`

```{r}
#| label: load-metadata
metadata <- read_csv(file.path(data_dir, "01_participant_metadata_period.csv"))
glimpse(metadata)
```

**What a single row represents:** one study participant, with their demographic
and clinical characteristics.

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | Participant ID. Primary key — each participant appears exactly once. |
| `arm` | character | Study arm: `birth_control` or `no_birth_control`. |
| `age` | numeric | Age of the participant in years (range 26–41). |
| `pcos_status` | character | Polycystic Ovary Syndrome status: `no disease` or `pcos`. |
| `period_product` | character | Menstrual product used: `tampon`, `pad`, or `menstrual_cup`. |
| `sex` | character | Sex assigned at birth. All entries are `female`. |

This table gives you **background info** about each participant that you can use
to group or colour-code your plots (e.g. by PCOS status or age).

---

## 3. `02_luminex_period.csv`

```{r}
#| label: load-luminex
luminex <- read_csv(file.path(data_dir, "02_luminex_period.csv"))
glimpse(luminex)
```

**What a single row represents:** the measured concentration of **one cytokine**
in **one sample**. The table is in **long (tidy) format** — each sample appears
in 8 rows (one per cytokine).

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | Which sample this measurement comes from. Links to `00_sample_ids_period.csv`. |
| `cytokine` | character | The name of the cytokine or chemokine being measured. Eight targets: `IL-1a`, `IL-1b`, `IL-6`, `TNFa`, `MIG`, `IFNg`, `IP-10`, `MIP-3a`. |
| `conc` | numeric | Concentration (pg/mL) as measured by the Luminex assay. |
| `limits` | character | Quality flag: `not_censored` (value is reliable), `below_detection_limit` (too low to measure accurately), or `above_detect_limit` (too high — the assay maxed out). |

> ⚠️ **Censored values:** When `limits` is `below_detection_limit`, the `conc`
> column contains the assay's **detection limit** for that cytokine (a fixed
> number per cytokine — e.g. 0.09 for IL-6, 0.24 for IL-1b, 1.53 for MIG).
> The true concentration is somewhere below this value. For
> `above_detect_limit` (only SAMP044's IL-1a at 2500 pg/mL), the true value is
> above the reported number. Decide how to handle censored data in your
> analysis (common approaches: treat as missing, keep the detection limit as-is,
> or impute).

```{r}
#| label: censoring-summary
luminex |>
  count(limits, cytokine) |>
  pivot_wider(names_from = limits, values_from = n, values_fill = 0)
```

You can use `pivot_wider()` to reshape this into a **wide** format with one
column per cytokine:

```{r}
#| label: pivot-luminex-wide
luminex_wide <- luminex |> 
  pivot_wider(names_from = cytokine, values_from = c(conc, limits))

luminex_wide
```

---

## 4. `03_flow_cytometry_period.csv`

```{r}
#| label: load-flow
flow <- read_csv(file.path(data_dir, "03_flow_cytometry_period.csv"))
glimpse(flow)
```

**What a single row represents:** the **cell counts** from one sample, after
flow cytometry and gating to identify different immune cell populations.

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | Which sample this comes from. Links to `00_sample_ids_period.csv`. |
| `live_cd19_negative` | numeric | Number of live cells **after excluding B cells** (CD19-negative). This is the starting population for all downstream gating. |
| `cd45_negative` | numeric | Cells that are **CD45-negative** (non-immune / stromal cells). |
| `cd45_positive` | numeric | Cells that are **CD45-positive** (immune cells / leukocytes). |
| `neutrophils` | numeric | Neutrophil count (a type of granulocyte, part of the innate immune system). |
| `non_neutrophils` | numeric | CD45-positive cells that are **not** neutrophils. |
| `cd3_negative` | numeric | Non-neutrophil cells that are **CD3-negative** (NK cells, monocytes, etc.). |
| `cd3_positive` | numeric | Non-neutrophil cells that are **CD3-positive** (all T cells). |
| `cd4_t_cells` | numeric | CD4⁺ T cells ("helper" T cells). |
| `cd8_t_cells` | numeric | CD8⁺ T cells ("killer" / cytotoxic T cells). |

The approximate gating strategy is:

```
Live + CD19- cells
  ├── CD45- (stromal / non-immune)
  └── CD45+ (leukocytes)
       ├── Neutrophils
       └── Non-neutrophils
            ├── CD3- (NK cells, monocytes, etc.)
            └── CD3+ (all T cells)
                 ├── CD4+ T cells  (helper)
                 ├── CD8+ T cells  (cytotoxic)
                 └── Other T cells (e.g. double-negative, γδ T cells)
```

> ⚠️ **The hierarchy is a logical gating tree, not a strict arithmetic
> one.** Some cells always fall through the gates (they express a marker
> at an unexpected level) or belong to a population not explicitly listed
> (e.g. double-negative CD4⁻CD8⁻ T cells, gamma-delta T cells).
> The code below shows where the gaps are.

```{r}
#| label: check-flow-additivity
# Check how well the gating hierarchy adds up (level by level)
flow |>
  mutate(
    # Level 0: starting population vs CD45- + CD45+
    l0_diff = live_cd19_negative - (cd45_negative + cd45_positive),
    # Level 1: CD45+ should ≈ neutrophils + non-neutrophils
    l1_diff = cd45_positive - (neutrophils + non_neutrophils),
    # Level 2: non-neutrophils should ≈ CD3- + CD3+
    l2_diff = non_neutrophils - (cd3_negative + cd3_positive),
    # Level 3: CD3+ should be >= CD4+ + CD8+ (it includes other subsets too)
    l3_gap = cd3_positive - (cd4_t_cells + cd8_t_cells)
  ) |>
  select(sample_id, l0_diff, l1_diff, l2_diff, l3_gap)
```

Key observations from the real data:

- **L0** (`live_cd19_negative` vs CD45⁻ + CD45⁺): Small differences exist
  (some events may fall outside the CD45 gate).
- **L1** (`cd45_positive` vs neutrophils + non-neutrophils): Very tight
  — most samples are within 10 cells of summing perfectly.
- **L2** (`non_neutrophils` vs CD3⁻ + CD3⁺): Always positive — about 900–2000
  cells on average fall outside the CD3 gate (possibly NK cells, dendritic
  cells, or monocytes with ambiguous CD3 staining).
- **L3** (`cd3_positive` vs CD4⁺ + CD8⁺): **Always positive** — CD3⁺ cells
  include not just CD4⁺ and CD8⁺ single-positive T cells, but also
  **double-negative** (CD4⁻CD8⁻) and **gamma-delta** T cells, and possibly
  **double-positive** (CD4⁺CD8⁺) immature T cells.

---

## Relationships between the tables

```
01_participant_metadata  ────  pid  ────  00_sample_ids_period
                                              │
                                              │ sample_id
                                              │
                    ┌─────────────────────────┼─────────────────────────┐
                    │ sample_id                                       │ sample_id
                    ▼                                                 ▼
          02_luminex_period                              03_flow_cytometry_period
```

- **`00_sample_ids_period`** is the **hub** — it links the other tables.
- Each `pid` in `01_participant_metadata` can appear many times in
  `00_sample_ids_period` (one row per sample collected).
- Each `sample_id` can appear once in `03_flow_cytometry_period` and 8 times
  (once per cytokine) in `02_luminex_period`.

### Joining everything together

```{r}
#| label: join-all
# Drop the duplicate "arm" from metadata (same as sample_ids)
all_data <- sample_ids |>
  left_join(metadata |> select(-arm), by = "pid") |>
  left_join(luminex, by = "sample_id") |>
  left_join(flow, by = "sample_id")

glimpse(all_data)
```