---
title: "Module 2 — Intro to R and Quarto"
format: html
editor: source
execute:
  warning: false
  message: false
---

This is your workspace for the Module 2 worksheet. The full instructions live
on the lesson page — keep it open alongside this notebook. You will build up
this notebook one part at a time: fill in the empty chunk under each heading,
press **Run Cell** to see what happens, and write a sentence of your own
between the chunks. Render with **Preview** whenever the worksheet tells you to.

Always load the packages you need in a setup chunk at the top. Run this first,
every time you open the notebook:

```{r setup}
library(tidyverse)
library(here)
```

## Part 1 — Run a chunk, write a sentence, render

# + is treated as a plus
# * is mltiply
# ** is 2 to the power of 10
# BODMAS aplies to both final examples

```{r}
3 + 5 + 10
10 * (5+1)
2 **10
64 / 2 + 2
64 / (2+2)
```

## Part 2 — Read the data and look at it

# There are 132 rows and 6 columns
# Columns that are interesting are pid and time_point
```{r}
table_02 <- read_csv(here(
  "data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
  table_02
  glimpse(table_02)
```

## Part 3 — Pull out columns and summarise them

# The mean blood crp is 1.47 excluding N/A values due to using the na.rm = TRUE function
```{r}
table_02$nugent_score
mean(table_02$crp_blood, na.rm = TRUE)
crp_mean <- mean(table_02$crp_blood, na.rm = TRUE)
crp_mean
```

## Part 4 — Build something with Pi

# Pi successfully wrote the correct code, with column names existing within the data
```{r}
 df <- read.csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
aggregate(nugent_score ~ arm, data = df, FUN = mean)
```

## Capstone — make it your own

```{r}
# Summary statistics for ph
library(knitr)

table_02 %>%
  summarise(
    min    = min(ph, na.rm = TRUE),
    max    = max(ph, na.rm = TRUE),
    median = median(ph, na.rm = TRUE),
    mean   = mean(ph, na.rm = TRUE),
    sd     = sd(ph, na.rm = TRUE)
  ) %>%
  kable(digits = 2, caption = "pH summary statistics")
```

# 46 Visits contained high Nugent scores indicating bacterial vaginosis. Furthermore, these cases contain higher average pH levels, which makes sense biologically as a healthy vaginal microbiome is acidic, with pH levels rising when harmful bacteria invade.
```{r}
# How many visits have a nugent_score of 7 or higher?
table_02 %>%
  filter(nugent_score >= 7) %>%
  nrow()
```

```{r}
# What's the average pH for high nugent_score (>=7) vs low (<7)?
table_02 %>%
  mutate(nugent_group = ifelse(nugent_score >= 7, "High (≥7)", "Low (<7)")) %>%
  group_by(nugent_group) %>%
  summarise(
    n    = n(),
    mean_ph = mean(ph, na.rm = TRUE),
    sd_ph   = sd(ph, na.rm = TRUE)
  ) %>%
  kable(digits = 2, col.names = c("Nugent Group", "Count", "Mean pH", "SD"),
        caption = "Average pH by nugent_score group")
```
## Before you go

Pick one step today that confused you. Ask Pi to explain it, then push back:
*"why did you use X instead of Y?"* Notice when Pi sounds confident versus when
it is guessing.
