---
title: "Module 3 — Data wrangling"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)

table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Part 1 — Predict (no code)

Take a look at `table_02` (run the cell above first). Answer in plain words below — no code yet.

```{r}
table_02
```

**1. How many rows? What does one row represent?**

*Your answer:* There are 132 rows that represent participent id's within each time point

**2. After `table_02 |> filter(time_point == "baseline")`, how many rows? What does each row represent now?**

*Your answer:* There will still be a reduced number of rows as only baseline patients are filtered

**3. After `table_02 |> count(arm)`, how many rows? What does each row represent now?**

*Your answer:* There will be 2 rows that are divided into plecabo and treatment (66 each if divided equally)

## Part 2 — Build the pipeline

Each step adds one verb on top of the previous pipeline. Run after each step.

```{r}
# step 1: pipe table_02 into head()
head(table_02)
```

```{r}
# step 2: filter to the baseline time point
table_02 |> filter(time_point == "baseline")
```

```{r}
# step 3: add select() for pid, arm, ph, nugent_score, crp_blood
table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood)
```

```{r}
# step 4: add mutate() for high_ph = if_else(ph >= 4.5, "yes", "no")
table_02 |> filter(time_point == "baseline") |> mutate(high_ph = if_else(ph >= 4, "yes", "no")) |> select(pid, arm, ph, nugent_score, crp_blood, high_ph)
```

```{r}
# step 5: add arrange(desc(ph))
table_02 |> filter(time_point == "baseline") |> mutate(high_ph = if_else(ph >= 4, "yes", "no")) |> select(pid, arm, ph, nugent_score, crp_blood, high_ph) |> arrange(desc(ph))
```

```{r}
# step 6: assign the final pipeline to early_visit, then print it
early_visit <- NULL  # replace NULL with the pipeline
early_visit <- table_02 |> filter(time_point == "baseline") |> mutate(high_ph = if_else(ph >= 4, "yes", "no")) |> select(pid, arm, ph, nugent_score, crp_blood, high_ph) |> arrange(desc(ph))
```

```{r}
# step 7: separately, table_02 |> count(arm)
table_02 |> count(arm)
```

## Part 3 — Reach (optional)

Add a `nugent_category` column with `case_when()`:

- 0–3 → "normal"
- 4–6 → "intermediate"
- 7–10 → "BV"

Then run `count(arm, nugent_category)`.

```{r}
# reach: mutate(nugent_category = case_when(...)) then count(arm, nugent_category)
table_02 |> 
  mutate(nugent_category = case_when(
    nugent_score >= 0 & nugent_score <= 3 ~ "normal",
    nugent_score >= 4 & nugent_score <= 6 ~ "intermediate",
    nugent_score >= 7 & nugent_score <= 10 ~ "BV"
  )) |> 
  count(arm, nugent_category)
```
