---
title: "Module 6 — Demographics and Table 1"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)

table_01 <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)
```

## Part 1 — Predict (no code)

```{r}
table_01
```

For each non-pid column of `table_01` (`arm`, `smoker`, `age`, `education`, `sex`), predict in plain words below. Don't run any code yet.

**1. `arm`**

- How will `tbl_summary(by = arm)` summarise it? *Your answer:* `arm` is the **grouping variable** (`by = arm`), so it won't appear as its own row. Instead, the table splits every other column into two columns: **placebo** vs **treatment**.
- What test will `add_p()` pick? *Your answer:* `add_p()` skips the grouping variable (`arm`) — it has nothing to test *on* `arm` itself. It adds tests to the **other** columns instead: smoker, age, education, and sex.

> **💡 Tutor note on your original answer:** You had the right idea — the table *does* compare treatment vs placebo. But "each variable stat in terms of treatment vs placebo" describes the *effect* of the grouping, not how `arm` itself is summarised. Since `arm` is the `by` variable, it has **no row** in the table. And `add_p()` doesn't select *which variables* to test — it picks a *statistical test* (e.g., chi-square, t-test) for each column. See Q2–5 below for the actual tests.

**2. `smoker`**

- *Your answer (summary, test):* **Summary:** counts and percentages (`n (%)`) of smokers vs non-smokers in each arm. **Test:** chi-square test of independence — tests whether smoking rate differs between treatment and placebo.

**3. `age`**

- *Your answer (summary, test):* **Summary:** median (Q1, Q3) — the default for a continuous variable. **Test:** Wilcoxon rank-sum test (non-parametric, used by `gtsummary` by default when n < 50) — tests whether the age distributions differ between arms.

**4. `education`**

- *Your answer (summary, test):* **Summary:** counts and percentages of each education level in each arm. **Test:** chi-square test of independence — tests whether education level is associated with arm assignment.

**5. `sex`**

- *Your answer (summary, test):* **Summary:** counts and percentages — but **all 44 participants are female**, so both arms show `F 100%`. **Test:** `add_p()` will **error** on sex because chi-square (and any test) needs at least two levels to compare, but sex has only one value (`"F"`).

## Part 2 — Build

Each step adds one piece on top of the previous table. Run after each step.

```{r}
# step 1: tbl_summary defaults
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm)
```

```{r}
# step 2: add p-values
# (modify the previous chunk by adding `|> add_p()`)
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p()
```

```{r}
# step 3: override age to mean (SD) + t-test
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test"))
```

```{r}
# step 4: polish
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test")) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

## Part 3 — Reach (optional)

```{r}
# reach: same table, but by = smoker instead of by = arm
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(
    by = smoker,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test")) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

> **What question is this table now answering?**
> It answers: **"How do participants' characteristics (age, education, arm assignment) differ between smokers and non-smokers?"**
>
> Instead of asking *"Did randomization work?"* (comparing treatment vs placebo), this asks a descriptive question: *"What does the smoking profile of our sample look like?"* — are smokers older? more educated? more likely to be in one arm?
